---
title: "E-commerce Revenue Forecasting using ARIMA and LSTM"
author: "Tran Nguyen Thanh Nam"
date: "`r Sys.Date()`"

output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6
)
```

# Business Understanding

Accurate revenue forecasting is critical for inventory planning, marketing budget allocation, and operational decision-making in e-commerce businesses.

This project analyzes the UCI Online Retail dataset and compares a statistical forecasting approach (ARIMA) with a deep learning approach (LSTM) for predicting future revenue trends.

# Project Objectives

- Clean and transform transaction-level data.
- Aggregate revenue into weekly_revenue and monthly_revenue time series.
- Explore trends and seasonality patterns.
- Develop ARIMA and LSTM forecasting models.
- Compare forecasting performance using MAE, RMSE, and MAPE.
- Generate revenue forecasts for the next 8 weeks.

0. Cài đặt & nạp thư viện
```{r}
pkgs <- c("readr", "dplyr", "lubridate", "ggplot2", "forecast",
          "tseries", "MASS", "keras", "zoo", "gridExtra")
new_pkgs <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(new_pkgs)) install.packages(new_pkgs)

library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(MASS)
library(keras)
library(zoo)
library(gridExtra)

dir.create("../images", showWarnings = FALSE)
dir.create("../reports", showWarnings = FALSE)
```

BƯỚC 1: NẠP & TIỀN XỬ LÝ DỮ LIỆU
```{r}

retail_data <- read_csv(
  "../data/online_retail.csv",
  locale = locale(encoding = "latin1")
)

# Làm sạch
retail_data <- retail_data %>%
  mutate(
    InvoiceDate = parse_date_time(InvoiceDate, orders = c("dmy HM", "ymd HM")),
    Revenue     = Quantity * UnitPrice
  ) %>%
  filter(Quantity > 0, UnitPrice > 0, !is.na(InvoiceDate))

cat("Dữ liệu sau làm sạch:", nrow(retail_data), "dòng\n")
cat("Khoảng thời gian:", format(min(retail_data$InvoiceDate)), "→", format(max(retail_data$InvoiceDate)), "\n")
cat("Tổng doanh thu: £", formatC(sum(retail_data$Revenue), format="f", digits=0, big.mark=","), "\n\n")

# ── Chuỗi doanh thu theo tuần ─────────────────────────────────
weekly_revenue <- retail_data %>%
  mutate(Week = floor_date(InvoiceDate, unit = "week")) %>%
  group_by(Week) %>%
  summarise(Revenue = sum(Revenue), .groups = "drop") %>%
  arrange(Week)

# ── Chuỗi doanh thu theo tháng ───────────────────────────────
monthly_revenue <- retail_data %>%
  mutate(Month = floor_date(InvoiceDate, unit = "month")) %>%
  group_by(Month) %>%
  summarise(Revenue = sum(Revenue), .groups = "drop") %>%
  arrange(Month)

cat("Số tuần:", nrow(weekly_revenue), "| Số tháng:", nrow(monthly_revenue), "\n\n")
```
BƯỚC 2: PHÂN TÍCH KHÁM PHÁ (EDA)
```{r}
# Vẽ chuỗi doanh thu tuần
p1 <- ggplot(weekly_revenue, aes(x = Week, y = Revenue)) +
  geom_line(color = "#378ADD", linewidth = 0.8) +
  geom_point(color = "#378ADD", size = 1.5, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "#D85A30",
              linewidth = 0.8, linetype = "dashed") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Doanh thu tuần – Online Retail",
       subtitle = "Đường đỏ: xu hướng LOESS",
       x = NULL, y = "Doanh thu (GBP)") +
  theme_minimal(base_size = 12)

# Vẽ doanh thu tháng
p2 <- ggplot(monthly_revenue, aes(x = Month, y = Revenue)) +
  geom_col(fill = "#378ADD", alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#D85A30",
              linewidth = 0.8, linetype = "dashed") +
  scale_y_continuous(labels = scales::comma_format(prefix = "£")) +
  labs(title = "Doanh thu tháng",
       subtitle = "Đường đỏ: xu hướng LOESS",
       x = NULL, y = "Doanh thu (GBP)") +
  theme_minimal(base_size = 12)

grid.arrange(p1, p2, ncol = 1)
ggsave(
  "../images/revenue_trend.png",
  plot = p1,
  width = 10,
  height = 6,
  dpi = 300
)

ggsave(
  "../images/monthly_revenue.png",
  plot = p2,
  width = 10,
  height = 6,
  dpi = 300
)
```
BƯỚC 3: CHUẨN BỊ CHUỖI THỜI GIAN
```{r}
weekly_ts <- ts(weekly_revenue$Revenue, frequency = 52)

cat("=== KIỂM ĐỊNH ADF ===\n")

# ADF cho chuỗi gốc
adf_orig <- adf.test(weekly_ts)

cat(
  "Chuỗi gốc: ADF =",
  round(adf_orig$statistic, 4),
  "| p-value =",
  round(adf_orig$p.value, 4),
  ifelse(
    adf_orig$p.value > 0.05,
    "→ NON-STATIONARY\n",
    "→ STATIONARY\n"
  )
)

# Sai phân bậc 1
ts_diff <- diff(weekly_ts)

adf_diff <- adf.test(ts_diff)

cat(
  "Sau sai phân d=1: ADF =",
  round(adf_diff$statistic, 4),
  "| p-value =",
  round(adf_diff$p.value, 4),
  ifelse(
    adf_diff$p.value < 0.05,
    "→ STATIONARY ✓\n\n",
    "→ CẦN DIFF THÊM\n\n"
  )
)

# ACF & PACF
png(
  "../images/stationarity_check.png",
  width = 1200,
  height = 1000
)

par(mfrow = c(2,2))

acf(
  weekly_ts,
  lag.max = 30,
  main = "ACF - Original Series"
)

pacf(
  weekly_ts,
  lag.max = 30,
  main = "PACF - Original Series"
)

acf(
  ts_diff,
  lag.max = 30,
  main = "ACF - Differenced Series"
)

pacf(
  ts_diff,
  lag.max = 30,
  main = "PACF - Differenced Series"
)

par(mfrow = c(1,1))

dev.off()
```
BƯỚC 4: MÔ HÌNH ARIMA
```{r}
cat("FIT MÔ HÌNH ARIMA\n")

# 4a. Grid search AIC tốt nhất
best_aic   <- Inf
best_order <- c(1, 1, 1)

for (p in 0:3) {
  for (d in 0:1) {
    for (q in 0:3) {
      tryCatch({
        m   <- Arima(weekly_ts, order = c(p, d, q))
        if (m$aic < best_aic) {
          best_aic   <- m$aic
          best_order <- c(p, d, q)
        }
      }, error = function(e) NULL)
    }
  }
}

cat("Mô hình ARIMA tốt nhất: ARIMA(",
    paste(best_order, collapse = ","), ")  AIC =", round(best_aic, 2), "\n")

# 4b. Fit mô hình tốt nhất
model_arima <- Arima(weekly_ts, order = best_order)
print(summary(model_arima))

# 4c. Kiểm tra phần dư
cat("\n--- Phân tích phần dư ARIMA ---\n")
checkresiduals(model_arima)

# Ljung-Box
lb <- Box.test(resid(model_arima), lag = 20, type = "Ljung-Box")
cat("Ljung-Box p-value:", round(lb$p.value, 4),
    ifelse(lb$p.value > 0.05,
           "→ Phần dư là white noise ✓\n\n",
           "→ Còn tự tương quan trong phần dư\n\n"))

# 4d. Độ chính xác trên tập test (80/20) 
n_test    <- floor(length(weekly_ts) * 0.2)
ts_train  <- head(weekly_ts, length(weekly_ts) - n_test)
ts_test   <- tail(weekly_ts, n_test)

arima_fit  <- Arima(ts_train, order = best_order)
arima_pred <- forecast(arima_fit, h = n_test)$mean

arima_mae  <- mean(abs(ts_test - arima_pred))
arima_rmse <- sqrt(mean((ts_test - arima_pred)^2))
arima_mape <- mean(abs((ts_test - arima_pred) / ts_test)) * 100

cat("=== ĐỘ CHÍNH XÁC ARIMA (test set) ===\n")
cat("MAE  =", formatC(arima_mae,  format = "f", digits = 0, big.mark = ","), "\n")
cat("RMSE =", formatC(arima_rmse, format = "f", digits = 0, big.mark = ","), "\n")
cat("MAPE =", round(arima_mape, 2), "%\n\n")

# 4e. Dự báo 8 tuần tới
arima_full <- Arima(weekly_ts, order = best_order)
arima_fc   <- forecast(arima_full, h = 8)

p_arima <- autoplot(arima_fc) +
  scale_y_continuous(
    labels = scales::comma_format(prefix = "£")
  ) +
  labs(
    title = paste0(
      "Forecast ARIMA(",
      paste(best_order, collapse = ","),
      ")"
    )
  ) +
  theme_minimal()

p_arima

ggsave(
  "../images/arima_forecast.png",
  plot = p_arima,
  width = 10,
  height = 6,
  dpi = 300
)
```
BƯỚC 5: MÔ HÌNH LSTM (keras)
```{r}
weekly_ts <- ts(weekly_revenue$Revenue, frequency = 52)
cat("FIT MÔ HÌNH LSTM\n")

library(keras)
library(magrittr)

# 5a. Chuẩn hoá MinMax 
ts_vals  <- as.numeric(weekly_ts)
ts_min   <- min(ts_vals)
ts_max   <- max(ts_vals)
ts_scale <- (ts_vals - ts_min) / (ts_max - ts_min)

# 5b. Tạo sequences 
LOOK_BACK <- 8

make_sequences <- function(data, lb) {
  n <- length(data) - lb
  X <- matrix(0, nrow = n, ncol = lb)
  y <- numeric(n)
  for (i in seq_len(n)) {
    X[i, ] <- data[i:(i + lb - 1)]
    y[i]   <- data[i + lb]
  }
  list(X = X, y = y)
}

seqs    <- make_sequences(ts_scale, LOOK_BACK)
X_all   <- seqs$X
y_all   <- seqs$y
split_i <- floor(nrow(X_all) * 0.8)

X_train <- array(X_all[1:split_i, ], 
                 dim = c(split_i, LOOK_BACK, 1))
y_train <- y_all[1:split_i]

X_test  <- array(X_all[(split_i+1):nrow(X_all), ], 
                 dim = c(nrow(X_all) - split_i, LOOK_BACK, 1))
y_test  <- y_all[(split_i+1):length(y_all)]

# 5c. Xây dựng & huấn luyện mô hình
set.seed(42)

model_lstm <- keras_model_sequential() %>%
  layer_lstm(units = 64, return_sequences = TRUE,
             input_shape = c(LOOK_BACK, 1)) %>%
  layer_dropout(rate = 0.2) %>%
  layer_lstm(units = 32) %>%
  layer_dropout(rate = 0.2) %>%
  layer_dense(units = 1)

model_lstm %>% compile(
  optimizer = optimizer_adam(),
  loss      = "mean_squared_error"
)

summary(model_lstm)

history <- model_lstm %>% fit(
  X_train, y_train,
  epochs          = 50,   # giảm để chạy nhẹ hơn
  batch_size      = 4,
  validation_data = list(X_test, y_test),
  verbose         = 1
)

# Vẽ learning curve
png(
  "../images/lstm_learning_curve.png",
  width = 1200,
  height = 800
)

plot(history)

dev.off()

plot(history)

# 5d. Đánh giá trên tập test 
lstm_pred_scale <- model_lstm %>% predict(X_test) %>% as.vector()

lstm_pred   <- lstm_pred_scale * (ts_max - ts_min) + ts_min
lstm_actual <- y_test          * (ts_max - ts_min) + ts_min

lstm_mae  <- mean(abs(lstm_actual - lstm_pred))
lstm_rmse <- sqrt(mean((lstm_actual - lstm_pred)^2))
lstm_mape <- mean(abs((lstm_actual - lstm_pred) / lstm_actual)) * 100

cat("=== ĐỘ CHÍNH XÁC LSTM (test set) ===\n")
cat("MAE  =", formatC(lstm_mae,  format = "f", digits = 0, big.mark = ","), "\n")
cat("RMSE =", formatC(lstm_rmse, format = "f", digits = 0, big.mark = ","), "\n")
cat("MAPE =", round(lstm_mape, 2), "%\n\n")

# 5e. Dự báo 8 tuần tới
seq_fc      <- tail(ts_scale, LOOK_BACK)
lstm_future <- numeric(8)

for (i in 1:8) {
  x_in           <- array(seq_fc, dim = c(1, LOOK_BACK, 1))
  p              <- model_lstm %>% predict(x_in) %>% as.numeric()
  lstm_future[i] <- p
  seq_fc         <- c(seq_fc[-1], p)
}

lstm_fc <- lstm_future * (ts_max - ts_min) + ts_min

cat("Dự báo 8 tuần tới (GBP):\n")
print(round(lstm_fc, 0))
```
BƯỚC 6: SO SÁNH & VISUALISE
```{r}
cat("=== SO SÁNH MÔ HÌNH ===\n")
cat(sprintf("%-10s  MAE=%s   RMSE=%s   MAPE=%.2f%%\n",
            paste0("ARIMA(", paste(best_order, collapse=","), ")"),
            formatC(arima_mae,  format="f", digits=0, big.mark=","),
            formatC(arima_rmse, format="f", digits=0, big.mark=","),
            arima_mape))
cat(sprintf("%-10s  MAE=%s   RMSE=%s   MAPE=%.2f%%\n",
            "LSTM",
            formatC(lstm_mae,  format="f", digits=0, big.mark=","),
            formatC(lstm_rmse, format="f", digits=0, big.mark=","),
            lstm_mape))

# ── Vẽ Actual vs Predicted (test set) ────────────────────────
n_test_lstm <- length(lstm_pred)
idx_test    <- (length(ts_vals) - n_test + 1):length(ts_vals)
idx_lstm    <- tail(idx_test, n_test_lstm)

retail_data_compare <- data.frame(
  idx    = c(idx_test, idx_lstm, idx_lstm),
  val    = c(ts_test, as.numeric(arima_pred), lstm_pred),
  model  = c(rep("Thực tế",   n_test),
             rep("ARIMA",     n_test),
             rep("LSTM",      n_test_lstm))
)

p_compare <- ggplot(
  retail_data_compare,
  aes(
    x = idx,
    y = val,
    color = model,
    linetype = model
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c(
      "Thực tế" = "#888780",
      "ARIMA"   = "#378ADD",
      "LSTM"    = "#1D9E75"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "Thực tế" = "solid",
      "ARIMA"   = "dashed",
      "LSTM"    = "dotdash"
    )
  ) +
  scale_y_continuous(
    labels = scales::comma_format(prefix = "£")
  ) +
  labs(
    title = "ARIMA vs LSTM – Test Set",
    x = "Week Index",
    y = "Revenue (GBP)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom"
  )

p_compare

ggsave(
  filename = "../images/model_comparison.png",
  plot = p_compare,
  width = 10,
  height = 6,
  dpi = 300
)

# ── Vẽ dự báo 8 tuần tới ─────────────────────────────────────
n_hist <- 12
hist_idx <- tail(seq_along(ts_vals), n_hist)
fc_idx   <- (length(ts_vals) + 1):(length(ts_vals) + 8)

retail_data_fc <- data.frame(
  idx   = c(hist_idx, fc_idx, fc_idx),
  val   = c(tail(ts_vals, n_hist), as.numeric(arima_fc$mean), lstm_fc),
  type  = c(rep("Lịch sử",   n_hist),
            rep("ARIMA FC",  8),
            rep("LSTM FC",   8))
)

# Khoảng tin cậy ARIMA
retail_data_ci <- data.frame(
  idx  = fc_idx,
  lo   = as.numeric(arima_fc$lower[, 2]),
  hi   = as.numeric(arima_fc$upper[, 2])
)

p_forecast <- ggplot() +

  geom_ribbon(
    data = retail_data_ci,
    aes(
      x = idx,
      ymin = lo,
      ymax = hi
    ),
    fill = "#378ADD",
    alpha = 0.15
  ) +

  geom_line(
    data = retail_data_fc,
    aes(
      x = idx,
      y = val,
      color = type,
      linetype = type
    ),
    linewidth = 1
  ) +

  geom_point(
    data = retail_data_fc,
    aes(
      x = idx,
      y = val,
      color = type
    ),
    size = 2.5
  ) +

  scale_color_manual(
    values = c(
      "Lịch sử"  = "#888780",
      "ARIMA FC" = "#378ADD",
      "LSTM FC"  = "#1D9E75"
    )
  ) +

  scale_linetype_manual(
    values = c(
      "Lịch sử"  = "solid",
      "ARIMA FC" = "dashed",
      "LSTM FC"  = "dotdash"
    )
  ) +

  scale_y_continuous(
    labels = scales::comma_format(prefix = "£")
  ) +

  labs(
    title = "Revenue Forecast for Next 8 Weeks",
    subtitle = "Blue area represents 95% confidence interval of ARIMA",
    x = "Week Index",
    y = "Revenue (GBP)"
  ) +

  theme_minimal(base_size = 12) +

  theme(
    legend.position = "bottom"
  )

p_forecast

ggsave(
  filename = "../images/forecast_8weeks.png",
  plot = p_forecast,
  width = 10,
  height = 6,
  dpi = 300
)
```
BƯỚC 7: IN KẾT QUẢ DỰ BÁO
```{r}
last_date <- max(weekly_revenue$Week)
fc_dates  <- last_date + weeks(1:8)

cat("\n=== DỰ BÁO 8 TUẦN TỚI ===\n")
cat(sprintf("%-15s  %12s  %12s  %12s  %12s\n",
            "Tuần", "ARIMA", "CI Lo 95%", "CI Hi 95%", "LSTM"))
cat(strrep("-", 70), "\n")
for (i in 1:8) {
  cat(sprintf("%-15s  %12s  %12s  %12s  %12s\n",
              format(fc_dates[i], "%d/%m/%Y"),
              paste0("£", formatC(arima_fc$mean[i],   format="f", digits=0, big.mark=",")),
              paste0("£", formatC(arima_fc$lower[i,2],format="f", digits=0, big.mark=",")),
              paste0("£", formatC(arima_fc$upper[i,2],format="f", digits=0, big.mark=",")),
              paste0("£", formatC(lstm_fc[i],          format="f", digits=0, big.mark=","))))
}

cat("\n=== HOÀN THÀNH ===\n")
```

# Conclusion

This project developed and evaluated ARIMA and LSTM models for e-commerce revenue forecasting.

Key contributions:
- End-to-end data cleaning and preprocessing pipeline.
- Statistical validation using Augmented Dickey-Fuller (ADF) stationarity testing.
- Comparison of traditional and deep learning forecasting methods.
- Practical business-oriented revenue forecasts for future planning.

Future improvements may include Prophet, XGBoost, external economic variables, holiday effects, and deployment through a Shiny dashboard.
